---
title: "Group project6_DA"
author: "soultan"
echo: false
eval: true

number-sections: true
format: html
editor: visual
---

```{r message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidymodels)


library(kableExtra)
library(gt)
library(tidyverse)

```

# Explorotary data analysis {#sec-EDA}

```{r}
mydata<-read.csv("c:\\DAProject6.csv")
```

# 

### **Analyzing BMI Distribution by Different Factors**

##### **Age**

```{r label=fig-line,fig.cap="The linear relation ship between Age and BMI",warning=FALSE,message=FALSE}
ggplot(mydata, aes(x = Age, y = BMI)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",se=FALSE, col = "blue") +
  theme_minimal() +
  labs(title = "BMI vs Age", x = "Age", y = "BMI")


```

From @fig-line we can see that there is a weak linear relation between age and BMI.

This weak relation tells that there is a small negative effect of the `Age` on the `BMI`.

##### Gender

```{r label=fig-sex,fig.cap="The boxplot of the Sex VS BMI",warning=FALSE,message=FALSE}
ggplot(mydata, aes(x = Sex, y = BMI, fill = Sex)) +
 geom_boxplot() +
  theme_minimal() +
  labs(title = "BMI Distribution by Gender", x = "Gender", y = "BMI")

```

From @fig-sex we can see that greater variability in BMI among females compared to males,however having the same median indicates that while central tendency is similar, BMI distribution differs between genders.

##### Education

```{r label=fig-education,fig.cap="BMI Distribution by Education",warning=FALSE,message=FALSE}
ggplot(mydata, aes(x = Education, y = BMI, fill = Education)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "BMI Distribution by Education", x = "Education Level", y = "BMI") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

From @fig-education we can see that the small different in medians suggest a minor impact of education on BMI. However the variation on boxplots spreads,indicates that higher education levels might be associated with lower BMI variability.

##### Veg

```{r warning=FALSE,message=FALSE,label=fig-veg}
plot1<-ggplot(mydata, aes(x = Veg, y = BMI, fill = Veg)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "BMI vs Vegetable Consumption", x = "Vegetable Intake", y = "BMI")

plot2<-ggplot(mydata, aes(x = Fruit, y = BMI, fill = Fruit)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "BMI vs Fruit Consumption", x = "Fruit Intake", y = "BMI")
library(patchwork)
plot1|plot2

```

From @fig-veg since both fruit and vegetarian have same boxplots distribution and median,it suggest that neither of them significantly effect BMI distribution.This indicates that BMI is similar regardless of whether a person having a fruit or vegetable.
