---
title: "Report on Body mass index(BMI)"
authors: "William Lang,  Freya Lynn, Abbey O'Donnell, Soultan Albarghouth"
       
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidymodels)
library(sjPlot)
library(performance)
library(kableExtra)
library(gt)
library(tidyverse)
library(moderndive)
library(skimr)
library(gapminder)
```

# Introduction {#sec-intro}

Body mass index(BMI) is an important variable in determing how healthy an indivdual is and is affected by such variables as diet and exercise. The following report will through stastical analysis answer the questions of intrest:

1.  Has the body mass index (BMI) in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any differences in the BMI distribution by age, gender, socio-economic status or lifestyle factors?

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| label: tbl-summmary
#| tbl-cap:  Mean, median and standard deviation (sd) of BMI by year.

library(gt)
library(dplyr)
bmi.data <- read.csv("DAProject6.csv")
bmi.data.year <- bmi.data |>
  select(BMI, Year)


```

| Year | Mean  | Median | Standard deviation |
|------|-------|--------|--------------------|
| 2016 | 28.03 | 27.28  | 5.79               |
| 2015 | 28.02 | 27.26  | 5.53               |
| 2014 | 27.84 | 27.17  | 5.21               |
| 2013 | 27.81 | 27.08  | 5.28               |

: Mean,Median and standard deviation of BMI by year

Therefore we can clearly see that the mean and median BMI has slightly increased over the years from 2013-2016. We can also see that the variation has also slightly increased from 2014-2016.

```{r}
bmi.year <- bmi.data.year
bmi.year$Year <- as.factor(bmi.year$Year)


ggplot(bmi.year, aes(y = BMI, x = Year, fill = Year))+
  geom_boxplot() +
  labs(x = "Year", y = "BMI", title = "BMI over the different years") +
  scale_fill_discrete(guide = "none")
```

We can therefore see from the boxplots that there is a slight increase in BMI over the years and that the normality assumption is not vaild due to the number of outliers in the data.


# Formal Analysis {#sec-FA}

## Question 1

To determine whether BMI has changed in Scotland over the given years, we fit a regression model, where BMI is our outcome variable y and Year is our categorical variable x :

```{r}
data <- read.csv("DAProject6.csv")

data$Year <- as.factor(data$Year)
  
model1 <- linear_reg() |>
  fit(BMI ~ Year, data = data)
```

We can use the estimates obtained here to fit our regression equation as follows:

$$
BMI = \alpha + \beta_{2014}\cdot I_{2014}(x) +\beta_{2015}\cdot I_{2015}(x)+\beta_{2016}\cdot I_{2016}(x)
$$

where

-   the intercept $\alpha$ is the mean BMI for our baseline year, 2013;

-   $\beta_{year}$ is the difference in the mean BMI in a given year relative to the baseline year 2013; and

-   $I_{year}(x)$ is an indicator function such that

    $$
    I_{year}(x) = \left\{ \begin{array}{rcl}\
    1 & \mbox{if x matches the given year}
     \\ 0 & \mbox{otherwise} 
    \end{array}\right.
    $$

We can now assess the fit of our model.

```{r}

regression_points <- get_regression_points(model1$fit)

```

We first do this by obtaining residuals. These are then plotted by year to test our assumptions:

```{r}
#| label: fig-reg-scatter
#| fig-cap: Scatterplot of Residuals by Year

ggplot(regression_points, aes(x = Year, y = residual)) +
  geom_jitter(width = 0.1) + 
  labs(x = "Year", y = "Residual") +
  geom_hline(yintercept = 0, col = "blue")

```

@fig-reg-scatter shows that the residuals are evenly spread and have mean zero. Therefore our assumption of constant variance appears to be valid.

We next plot a histogram to check that the residuals are normally distributed:

```{r}
#| label: fig-reg-hist 
#| fig-cap: Histogram of Residuals
ggplot(regression_points, aes(x = residual)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(x = "Residual")
```

@fig-reg-hist shows that the residuals are centered at zero and are bell shaped. Therefore our assumption of normality appears to be valid.
## Question 2

```{r}
bmi <- read_csv("~/Downloads/DAProject6.csv", show_col_types = FALSE)

library(ggplot2)
library(dplyr)
library(skimr)
ggplot(bmi, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "blue4", color = "black", alpha = 0.5) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Distribution of BMI",
       x = "BMI",
       y = "Frequency") +
  theme_minimal()


```
```{r}
lm_spec <- linear_reg() |> set_engine("lm")
bmi.model1 <- lm_spec |> fit(BMI ~ Age + Sex + Education + Veg + Fruit +Year, data = bmi)
bmi.model1 <- bmi.model1 |> extract_fit_engine()
get_regression_table(bmi.model1)
```

```{r}
lm_spec <- linear_reg() 
bmi.model1.int <- lm_spec |> fit(BMI ~ Age * Sex * Education * Veg * Fruit * Year, data = bmi)
bmi.model1.int <- bmi.model1.int |> extract_fit_engine()
get_regression_table(bmi.model1.int)
```
```{r}

library(tidymodels)
library(moderndive) 
library(broom)      
library(dplyr)

lm_spec <- linear_reg()

bmi_full_model <- lm_spec %>% 
  fit(BMI ~ Age + Sex + Education + Veg + Fruit + Year, data = bmi)


bmi_full_model_fit <- bmi_full_model |> extract_fit_engine()
full_results <- tidy(bmi_full_model_fit, conf.int = TRUE)
print(full_results)


bmi_model_int <- lm_spec |>
  fit(BMI ~ Age + Sex + Education + Veg + Fruit + Year + 
        Age:Sex +           
        Sex:Education +     
        Veg:Fruit +         
        Year:Age,           
      data = bmi)


bmi_model_int_fit <- bmi_model_int |> extract_fit_engine()

int_results <- get_regression_table(bmi_model_int_fit)
print(int_results)


model_comparison <- bind_rows(
  tibble(model = "Full Model", 
         AIC = AIC(bmi_full_model_fit),
         BIC = BIC(bmi_full_model_fit)),
  tibble(model = "Interaction Model", 
         AIC = AIC(bmi_model_int_fit),
         BIC = BIC(bmi_model_int_fit))
)
print(model_comparison)


plot(bmi_model_int_fit)
```

# Conclusions {#sec-con}

