---
title: "Report on Body mass index(BMI)"
author: "William Lang"
execute: 
  echo: false
  eval: true
number-sections: true
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidymodels)
library(sjPlot)
library(performance)
library(kableExtra)
library(gt)
```

# Introduction {#sec-intro}

Body mass index(BMI) is an important variable in determing how healthy an indivdual is and is affected by such variables as diet and exercise. The following report will through stastical analysis answer the questions of intrest:

1.  Has the body mass index (BMI) in Scotland changed over the given years of the Scottish Health Survey?

2.  Are there any differences in the BMI distribution by age, gender, socio-economic status or lifestyle factors?

# Exploratory Analysis {#sec-EA}

```{r}
#| echo: false
#| label: tbl-summmary
#| tbl-cap:  Mean, median and standard deviation (sd) of BMI by year.

library(gt)
library(dplyr)
bmi.data <- read.csv("DAProject6.csv")
bmi.data.year <- bmi.data |>
  select(BMI, Year)
bmi.data.year |>
   summarize('BMI.Mean' = mean(BMI),
             'BMI.Median' = median(BMI),
             'BMI.sd' = sd(BMI),
             .by = Year) |>
 gt() |> 
  fmt_number(decimals=2) |>
  tab_spanner(
    label = "BMI",
    columns = c(BMI.Mean, BMI.Median,BMI.sd)
  ) |>
  cols_label(
    BMI.Mean = html("Mean"),
    BMI.Median = html("Median"),
    BMI.sd = html("Std.Dev"),
  )

```

| Year | Mean  | Median | Standard deviation |
|------|-------|--------|--------------------|
| 2016 | 28.03 | 27.28  | 5.79               |
| 2015 | 28.02 | 27.26  | 5.53               |
| 2014 | 27.84 | 27.17  | 5.21               |
| 2013 | 27.81 | 27.08  | 5.28               |

: Mean,Median and standard deviation of BMI by year

Therefore we can clearly see that the mean and median BMI has slightly increased over the years from 2013-2016. We can also see that the variation has also slightly increased from 2014-2016.

```{r}
bmi.year <- bmi.data.year
bmi.year$Year <- as.factor(bmi.year$Year)

glimpse(bmi.data.year)
glimpse(df2)


ggplot(bmi.year, aes(y = BMI, x = Year, fill = Year))+
  geom_boxplot() +
  labs(x = "Year", y = "BMI", title = "BMI over the different years") +
  scale_fill_discrete(guide = "none")
```

We can therefore see from the boxplots that there is a slight increase in BMI over the years and that the normality assumption is not vaild due to the number of outliers in the data.
# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}

# Further Questions {#sec-further}

## Further Question 1 {#sec-further1}

1.  Answer to question 1

```{r}
#| echo: true
#| eval: true
# Copy your code here


```

2.  Answer to question 2: Write your equation:

$$
y_i = \beta_0 + \beta_1 x_1 + ...
$$

## Further Question 2 {#sec-further2}

### Part 1

1.  Answer to question 1:

2.  Answer to question 2:

3.  Answer to question 3:

4.  Answer to question 4:

5.  Answer to question 5:

### Part 2
